Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation

### 2023-06-16 docs/papers/sto/so.tex
FIXUP draft: abstract and loss
FIXUP SO draft: rm a paragraph on TO

\begin{abstract}
We sharpen the PM force with first neural networks and then symbolic
regression, taking symmetry and dimensional analysis in consideration.
\end{abstract}

@@

\YL{
For a field $f$ in either Lagrangian or Eulerian space, we have thought
of 3 forms of MSE-inspired loss:
%
\begin{equation}
\cL_f = \ln \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
                   {\sum_{\vq'} f(\vq')^2},
\end{equation}
%
(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
their Fourier wavevector $\vk$ in both cases)
%
\begin{equation}
\cL_f = \frac1{N_K} \sum_i \ln
\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} | f(\vk')|^2},
\end{equation}
%
and
%
\begin{equation}
\cL_f = \ln \biggl[ \frac1{N_K} \sum_i
\frac{\sum_{k \in K_i} w(\vk)
      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
\end{equation}
%
Then the question is which loss we should use for which $f$, to get
stable and robust results, e.g., no bump for large or small boxes.
To further limit our scope in this paper, we should probably only
consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
velocity $\vv(\vq)$.
}

@@

We combine Lagrangian and Eulerian fields to form the total loss:
%
\begin{equation}
\cJ = \frac1{N_a} (\cL_\vs + [\cL_\vv] + \cL_\delta),
\end{equation}
%
where $N_a$ is the time steps it take to generate the snapshot for each
iteration.
\YL{(a simple sum since we have already done careful normalization in
$\cL_f$'s) and also try}
%
\begin{equation}
\cJ = \frac1{N_a} (\ln\cL_\vs + [\ln\cL_\vv] + \ln\cL_\delta).
\end{equation}
%

@@

\subsection{Temporal Optimization}
\label{sec:to}

We use a spline to model the function $a(t)$ that maps uniformly sampled $t$
values to optimally distributed $a$ time steps.
The spline is parameterized with a number of control points that are optimized
during training.
The gradients (i.e. cotangents) are back propagated from the array of
integration time steps to the control points through a VJP transformation of
$a(t)$.

The time-dependence of the discrete symplectic integrator mainly enters through
the drift and kick factors in \eqref{DFK}.
Besides, the input features of the SO NNs introduced in \autoref{sec:so} could
also depend on time.
Therefore when we perform TO together with SO, the force evaluation in
\eqref{DFK} which contains SO also becomes time-dependent and needs to be
optimized.

>>>>

\begin{abstract}
We sharpen the PM force with first neural networks and then symbolic
regression, taking symmetry and dimensional analysis in consideration.

We combine differentiable simulation, deep learning, [decision tree
(feature selection)], and symbolic regression to extract the final kernel
applicable to all other cosmological simulation code.
\end{abstract}

@@

\YL{
For a field $f$ in either Lagrangian or Eulerian space, we have thought
of 3 forms of MSE-inspired loss:
%
\begin{equation}
\cL_f = \frac1{N_a} \ln
  \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
       {\sum_{\vq'} f(\vq')^2},
\end{equation}
%
where $N_a$ is the time steps it take to generate the snapshot for each
iteration.
(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
their Fourier wavevector $\vk$ in both cases)
%
\begin{equation}
\cL_f = \frac1{N_a N_K} \sum_i \ln
\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} | f(\vk')|^2},
\end{equation}
%
and
%
\begin{equation}
\cL_f = \frac1{N_a} \ln \biggl[ \frac1{N_K} \sum_i
\frac{\sum_{k \in K_i} w(\vk)
      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
\end{equation}
%
Then the question is which loss we should use for which $f$, to get
stable and robust results, e.g., no bump for large or small boxes.
To further limit our scope in this paper, we should probably only
consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
velocity $\vv(\vq)$.
}

@@

We combine Lagrangian and Eulerian fields to form the total loss:
%
\begin{equation}
\cJ = \cL_\vs + [\cL_\vv] + \cL_\delta.
\end{equation}
%

@@
